### 2022-10-26 draft/draft.Rmd
minor tweak to proof of theorem 4.2
minor updates
adding in detaisls on bezier curve fitting

Graph and network data have become increasingly widespread in various fields including sociology, neuroscience, biostatistics, and computer science. 
This has resulted in challenges for researchers who rely on traditional statistical and machine learning methods that are incompatible with graph data and instead assume that the data exist as feature vectors. 
To illustrate this, consider the typical approach to building a statistical or machine learning model. 
Data are often represented as an $n \times p$ matrix $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$ in which each row $x_i \in \mathbb{R}^p$ is an observation of $p$ features and each column is a set of $n$ feature measurements. 
An analysis task for these data might be to come up with a classification model $\phi : \mathbb{R}^p \to \{1, 2, ..., K\}$ that uses the numerical values of each feature of a vector $x_i$ to calculate a predicted label $z_i \in \{1, 2, ..., K\}$. 
For instance, $\phi(x)$ might first compute the distances from $x$ to $K$ points in $\mathbb{R}^p$ and then assign $x$ to the label of the nearest point. 
Examples of this include linear discriminant analysis (in the case of supervised learning) and Lloyd's algorithm \citep{1056489} or Gaussian mixture models \citep{doi:10.1198/016214502760047131} (in the case of unsupervised learning). 
However, it is not obvious how this method would translate to data that are represented as graphs, in which observations consist of relationships among a set of objects rather than numerical attributes associated with each object: 
Instead of feature vector $x_i = \begin{bmatrix} x_{i1} & \cdots & x_{ip} \end{bmatrix}^\top \in \mathbb{R}^p$, we observe $a_i = \begin{bmatrix} a_{i1} & \cdots & a_{in} \end{bmatrix}^\top \in \mathbb{R}^n$ in which each $a_{ij}$ is object $i$'s relationship to object $j$. 
For such data, it is often necessary to alter existing algorithms, transform the graph data into Euclidean data, often called *graph embedding* or *spectral clustering* \citep{vonLuxburg2007}, and apply algorithms for Euclidean data on the embedding, or come up with new algorithms altogether.

@@

To estimate $\omega = \begin{bmatrix} \omega_1 & \cdots & \omega_n \end{bmatrix}^\top$ after observing $A$ and supposing that the relative locations of the planets as well as the normalizing constant $C$ are known, one approach might be maximum likelihood maximization. 
For simplicity, set $C = 1$. 
Then the log-likelihood function can be written as:
$$
\ell(\omega; A) = 
\sum_{i < j} A_{ij} \log \bigg( \frac{\omega_i \omega_j}{d_{ij}^2 -
\omega_i \omega_j} \bigg) + \log (d_{ij}^2 - \omega_i \omega_j) + const.
$$
The partial derivatives with respect to each $\omega_i$ are:
$$
\frac{\partial \ell}{\partial \omega_i} = \sum_{j \neq i} \frac{A_{ij}}{\omega_i} - \frac{\omega_j (1 - A_{ij})}{d_{ij}^2 - \omega_i \omega_j}.
$$
Setting the gradient to zero yields $n$ equations with $n$ unknowns, but it is not entirely obvious how to solve this system of equations. 
Alternatively, we can use gradient ascent or some other numerical optimization method, since this particular log-likelihood function is concave. 
\end{example}

@@

1. Let the size of the graph, $n$, be fixed. 
Then each vertex $v_i$, and its corresponding label, $z_i$, are also fixed, in addition to the pairwise community edge probabilities $\{\theta_{k \ell}\}_K$, which are all treated as model parameters. 
Under this sampling scheme, there is no notion of increasing $n$. 
2. The vertices and edges are sampled in sequence, and we suppose that each vertex $v_i$ comes with a corresponding label, $z_i$, which determines its edge probabilities to the other vertices. 
Under this sampling scheme, there is no assumed distribution on the labels, which are treated as fixed parameters rather than random variables, along with $\{\theta_{k \ell}\}_K$, but there is a notion of increasing sample size $n$, which allows for asymptotics. 
3. The vertices are sampled in sequence, and each label, $z_i$, is sampled as $z_1, ..., z_n \iid \Multinomial(\alpha_1, ..., \alpha_K)$, $\sum_k^K \alpha_k = 1$. 
The edge probabilities are then determined by the sample $\{z_i\}$ and the parameters $\{\theta_{k \ell}\}_K$.
Under this sampling scheme, both the edges and the labels are random, with fixed parameters $\alpha = \begin{bmatrix} \alpha_1 & \cdots & \alpha_K \end{bmatrix}^\top$ and $\{\theta_{k \ell}\}_K$.

@@

\begin{definition}[Generalized random dot product graph]
\label{def:grdpg}
Let $p \geq 1$ and $q \geq 0$ be integers. 
Define $I_{p,q}$ as the block diagonal matrix $I_{p,q} = \Bigl[\begin{smallmatrix} I_{p} & 0 \\ 0 & - I_{q} \end{smallmatrix} \Bigr]$ where $I_p$ and $I_q$ are the identity matrices of dimensions $p \times p$ and $q \times q$ respectively. 
Denote $d = p + q$ and let $\mathcal{X}$ be the subset of $\mathbb{R}^d$ such that, for any $x, y \in \mathcal{X}$, we have $x^\top I_{p,q} y \in [0, 1]$. 
Let $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$ be an $n \times d$ matrix with rows $x_i \in \mathcal{X}$. 
A graph $G$ with adjacency matrix $A$ is said to be a generalized random dot product graph with latent positions $X$, sparsity parameter $\rho_n$, and signature $(p, q)$ if $A \sim \BG(P)$ where the edge probability matrix $P$ is given by $P = \rho_n X I_{p,q} X^\top$, i.e., the entries of $P$ are of the form $P_{ij} = \rho_n x_i^\top I_{p,q} x_j$.

@@

\begin{example}
Suppose that latent vectors for an RDPG are drawn from the curve 
$x(t) = \begin{bmatrix} t^2 & 2 t (t-1) \end{bmatrix}^\top$, $t \in [0, 1]$. 
Let $x_i = x(t_i)$. 
Then for each $i < j$, the edges are drawn as $A_{ij} \indep \Bernoulli(x_i^\top x_j)$. 
If after observing $A$, we wish to estimate the curve, the ASE provides an estimator in which the estimated latent vectors converge to the curve, up to an orthogonal transformation (figure \ref{fig:ase-example}).

@@

\begin{example}
In the $\ER$ model, the edge probability matrix has a fixed value $P_{ij} \equiv p \in [0, 1]$. 
This corresponds to a one-dimensional RDPG with a latent structure $X = \begin{bmatrix} \sqrt{p} & \cdots & \sqrt{p} \end{bmatrix}^\top$. 
In another view, the RDPG with latent structure $X = \begin{bmatrix} \sqrt{p} & \cdots & \sqrt{p} \end{bmatrix}^\top$ induces an $\ER$ model in which the edge probabilities have a fixed value.

Suppose that we have RDPG in which the latent space is in $\mathbb{R}^2$ and the latent vectors are drawn uniformly from the quarter unit circle defined by $g(t) = \begin{bmatrix} \cos \frac{\pi}{2} t & \sin \frac{\pi}{2} t \end{bmatrix}^\top$ for $t \in [0, 1]$. 
Then it can be shown that instead of a fixed $P_{ij} = p$, this RDPG induces a model in which the edge probabilities follow a distribution with density $f(p) = \frac{2}{\pi - 2} \Big( \frac{1}{1 - p^2} - 1 \Big)$. 
\end{example}

@@

\begin{enumerate}
\item Draw labels $z_1, ..., z_n \iid \Multinomial(\alpha_1, ..., \alpha_K)$.
\item Draw latent vectors by first taking each $t_i \indep F_{z_i}$ and then computing each $x_i = g_{z_i}(t_i)$. 
\item Compile the latent vectors into data matrix $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$ and define the adjacency matrix as $A \sim \GRDPG_{p,q}(X; \rho_n)$. 
\end{enumerate}
\end{definition}

@@

\begin{example}
\label{ex:separated-curves}
Define two one-dimensional manifolds in $\mathbb{R}^2$ by $f_1(t) = \begin{bmatrix} \cos(\frac{\pi}{3} t) & \sin(\frac{\pi}{3} t) \end{bmatrix}^\top$ and $f_2(t) = \begin{bmatrix} 1 - \cos(\frac{\pi}{3} t) & 1 - \sin(\frac{\pi}{3} t) \end{bmatrix}^\top$.
Draw $t_1, ..., t_n \iid \Uniform(0, 1)$ and $z_1, ..., z_n \iid \Multinomial(\frac{1}{2}, \frac{1}{2})$, and compute latent vectors $x_i = f_{z_i}(t_i)$, which are collected in data matrix $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$. 
Finally, let $A \sim \RDPG(X)$. Figure~\ref{fig:two-curves} shows the latent configuration drawn from this latent distribution, a random dot product graph drawn from the latent configuration, and the ASE of the graph. 
Although the community structure is not obvious from the graph, the embedding shows a clear separation between the two communities.

@@

\begin{theorem}[Community detection for the GRDPG for which the communities come from nonintersecting manifolds]
\label{theorem:nonintersect-grdpg}
Let $x_1, ..., x_n$ be points sampled on $K$ compact, connected manifolds $\mathcal{M}_1, ..., \mathcal{M}_K \subset \mathbb{R}^d$ each with probability measures $F_1, ..., F_K$, and the manifolds are separated by distance at least $\delta = \min\limits_{k \neq \ell} \min\limits_{x_i \in \mathcal{M}_k, x_j \in \mathcal{M}_\ell} \| x_i - x_j \| > 0$. 
Let $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$ and $A \sim \GRDPG_{p,q}(X; \rho_n)$ for some $p, q \in \mathbb{N}_0$ such that $p + q = d$ and sparsity parameter $\rho_n$ that satisfies $n \rho_n = \omega(\log^c n)$ for some $c > 1$. 
Define $A_n(\eta)$ as the event that an $\eta$-neighborhood graph constructed from the ASE of $A$ consists of exactly $K$ disconnected subgraphs that map exactly to each manifold. 
Then for some $C > 0$ and any $\eta \in (0, C \delta)$, $\lim\limits_{n \to \infty} P(A_n(\eta)) = 1$. 
\end{theorem}

@@

1. Draw $t_1, ..., t_n \iid F$ for probability distribution $F$ with support $[0, 1]$.
2. Draw $z_1, ..., z_n \iid \Multinomial(\alpha_1, ..., \alpha_K)$, the community labels.
3. Let each $x_i = g_{z_i}(t_i)$ be the latent vector for vertex $v_i$, and collect the latent vectors into matrix $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$.
4. Draw $A \sim \GRDPG_{p,q}(X; \rho_n)$.

\begin{example}
\label{ex:intersect-curves}
Suppose we have a two-dimensional RDPG in which latent vectors lie on one of two curves defined by the parameterizations $g_1(t) = \begin{bmatrix} t^2 & 2 t (1-t) \end{bmatrix}^\top$ and $g_2(t) = \begin{bmatrix} 2 t (1-t) & (1-t)^2 \end{bmatrix}^\top$. 
This can be described as a two-community MBM in which the manifolds are both one-dimensional. 
For simplicity, we restrict $F_1 = F_2 = \Uniform(0, 1)$. 
$g_1$ and $g_2$ intersect at two points, $t = 0$ and $t = 2/3$, which in the latent space correspond to $(0, 0)$ and $(4/9, 4/9)$. 
Clearly, single-linkage or $\eta$-neighborhood clustering does not work for this setting, both in the original latent space and in the ASE. 
The community structure is also not clear from the adjacency matrix. 
However, there still is a clear pattern in the ASE that reveals the community structure (figure \ref{fig:intersect-curves-example}).

@@

\begin{example}
\label{ex:intersect-curves-fit}
Figure \ref{fig:intersect-curves-2} shows one result of applying algorithm \ref{alg:kcurves} on example \ref{ex:intersect-curves}. 
In this implementation, $g_1$ and $g_2$ are assumed to be Bezier curves in which $p_1^{(0)} = p_2^{(0)} = \begin{bmatrix} 0 & 0 \end{bmatrix}^\top$, i.e., the curves meet at the origin. 
Here, an accuracy of 87.5\% is achieved.

@@

For a simulation study, we again turn to examples.  \ref{ex:intersect-curves} and \ref{ex:intersect-curves-fit}. 
The exact simulation setup is as follows:

1. Draw $z_1, ..., z_n \iid \Multinomial(1/2, 1/2)$.
2. Draw $t_1, ..., t_n \iid \Uniform(0, 1)$. 
3. Let each $x_i = g_{z_i}(t_i)$ where  $g_1(t) = \begin{bmatrix} t^2 & 2 t (1-t) \end{bmatrix}^\top$ and $g_2(t) = \begin{bmatrix} 2 t (1-t) & (1-t)^2 \end{bmatrix}^\top$. 
Collect the latent vectors into matrix $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$. 
4. Draw $A \sim \RDPG(X; \rho_n)$ (for these simulations we fix $\rho_n \equiv 1$).

For this simulation study, we set $n = 128, 256, 512, 1024, 2048$, and we start with completely random initialization as well as initializations in which we know 4 or 8 labels from each community. 
For each setting, we simulate 50 replicates. 
Since the curves meet at the origin, we force the fitted Bezier curves to meet at the origin as well by fixing $p_1^{(0)} = p_2^{(0)} = (0, 0)$. 
Figure \ref{fig:sim-curves} shows the median and interquartile ranges for the error rates in each simulation setting. 
The simulation results suggest that for large $n$, we do not benefit from having more known labels, but having a few results in greatly improved performance.

```{r sim-curves, fig.height = 3, fig.width = 8, fig.cap = 'Median and IQR error rates over 50 replicates.', cache = TRUE}
n.vec <- 2 ^ c(7, 8, 9, 10, 11)
nsamp.vec <- c(0, 4, 8)

clustering.df <- readr::read_csv('~/dev/manifold-block-models/simulations/balanced.csv')

@@

@@

plot.estimated.curves(Xhat, mbconnectome.out) + 
  scale_colour_brewer(palette = 'Set1') + 
  coord_fixed(.5) +
  labs(x = expression(x[1]), y = expression(x[2]))
```

@@

Since $x_i = g(t_i; p_{z_i})$, $\|Q \hat{x}_i - g(t_i; p_{z_i})\| < \epsilon$.

The statement of the theorem assumes that for each community $k$, there at least $R + 1$ points with known label $k$. 
For simplicity, suppose we have exactly $R + 1$ (if there are more, then discard the rest). 
Then for each community $k$, it is possible to fit a polynomial of order $R$ that goes through each of the $R + 1$ known points exactly (see appendix C), yielding $g(t; \hat{p}_k)$, where $\hat{p}_k$ are the fitted coefficients of the Bezier polynomial. 
Use the fitted Bezier curves to define $\delta_k(t) = \|Q g(t; \hat{p}_k) - g(t; p_k)\|$, $\delta_k = \max_t \delta_k(t)$, and $\delta = \max_k \delta_k$. 
Since $P(\epsilon < M n^{-1/2} \log^c n) \to 1$ for constants $M, c > 0$, 
$P(\delta_k < M n^{-1/2} \log^c n) \to 1$ for each $k$. 
Since $\epsilon_i$ describes the error between each embedding point and its corresponding latent vertex, and $\delta_k(t_i)$ describes the error between the fitted curve and its corresponding true curve in the latent space, 
$$L < \frac{1}{n} \sum_i \epsilon_i + \delta_{z_i}(t_i) < \frac{1}{n} n (\epsilon + \delta) = \epsilon + \delta.$$ 
Then as $n \to \infty$, 
$$P(L < 2 M n^{-1/2} \log^c n) > P(\epsilon + \delta < 2 M n^{-1/2} \log^c n) \to 1.$$

\end{proof}

@@

>>>>

Graph and network data have become increasingly widespread in various fields including sociology, neuroscience, biostatistics, and computer science. 
This has resulted in challenges for researchers who rely on traditional statistical and machine learning methods that are incompatible with graph data and instead assume that the data exist as feature vectors. 
To illustrate this, consider the typical approach to building a statistical or machine learning model. 
Data are often represented as an $n \times p$ matrix $X = \big[ x_1 \mid \cdots \mid x_n \big]^\top$ in which each row $x_i \in \mathbb{R}^p$ is an observation of $p$ features and each column is a set of $n$ feature measurements. 
An analysis task for these data might be to come up with a classification model $\phi : \mathbb{R}^p \to \{1, 2, ..., K\}$ that uses the numerical values of each feature of a vector $x_i$ to calculate a predicted label $z_i \in \{1, 2, ..., K\}$. 
For instance, $\phi(x)$ might first compute the distances from $x$ to $K$ points in $\mathbb{R}^p$ and then assign $x$ to the label of the nearest point. 
Examples of this include linear discriminant analysis (in the case of supervised learning) and Lloyd's algorithm \citep{1056489} or Gaussian mixture models \citep{doi:10.1198/016214502760047131} (in the case of unsupervised learning). 
However, it is not obvious how this method would translate to data that are represented as graphs, in which observations consist of relationships among a set of objects rather than numerical attributes associated with each object: 
Instead of feature vector $x_i = \big[ x_{i1} \mid \cdots \mid x_{ip} \big]^\top \in \mathbb{R}^p$, we observe $a_i = \big[ a_{i1} \mid \cdots \mid a_{in} \big]^\top \in \mathbb{R}^n$ in which each $a_{ij}$ is object $i$'s relationship to object $j$. 
For such data, it is often necessary to alter existing algorithms, transform the graph data into Euclidean data, often called *graph embedding* or *spectral clustering* \citep{vonLuxburg2007}, and apply algorithms for Euclidean data on the embedding, or come up with new algorithms altogether.

@@

To estimate $\omega = \big[ \omega_1, \cdots, \omega_n \big]^\top$ after observing $A$ and supposing that the relative locations of the planets as well as the normalizing constant $C$ are known, one approach might be maximum likelihood maximization. 
For simplicity, set $C = 1$. 
Then the log-likelihood function can be written as:
$$
\ell(\omega; A) = 
\sum_{i < j} A_{ij} \log \bigg( \frac{\omega_i \omega_j}{d_{ij}^2 -
\omega_i \omega_j} \bigg) + \log (d_{ij}^2 - \omega_i \omega_j) + const.
$$
The partial derivatives with respect to each $\omega_i$ are:
$$
\frac{\partial \ell}{\partial \omega_i} = \sum_{j \neq i} \frac{A_{ij}}{\omega_i} - \frac{\omega_j (1 - A_{ij})}{d_{ij}^2 - \omega_i \omega_j}.
$$
Setting the gradient to zero yields $n$ equations with $n$ unknowns, but it is not entirely obvious how to solve this system of equations. 
Alternatively, we can use gradient ascent or some other numerical optimization method, since this particular log-likelihood function is concave. 
\end{example}

@@

1. Let the size of the graph, $n$, be fixed. 
Then each vertex $v_i$, and its corresponding label, $z_i$, are also fixed, in addition to the pairwise community edge probabilities $\{\theta_{k \ell}\}_K$, which are all treated as model parameters. 
Under this sampling scheme, there is no notion of increasing $n$. 
2. The vertices and edges are sampled in sequence, and we suppose that each vertex $v_i$ comes with a corresponding label, $z_i$, which determines its edge probabilities to the other vertices. 
Under this sampling scheme, there is no assumed distribution on the labels, which are treated as fixed parameters rather than random variables, along with $\{\theta_{k \ell}\}_K$, but there is a notion of increasing sample size $n$, which allows for asymptotics. 
3. The vertices are sampled in sequence, and each label, $z_i$, is sampled as $z_1, ..., z_n \iid \Multinomial(\alpha_1, ..., \alpha_K)$, $\sum_k^K \alpha_k = 1$. 
The edge probabilities are then determined by the sample $\{z_i\}$ and the parameters $\{\theta_{k \ell}\}_K$.
Under this sampling scheme, both the edges and the labels are random, with fixed parameters $\alpha = \big[ \alpha_1, \cdots, \alpha_K \big]^\top$ and $\{\theta_{k \ell}\}_K$.

@@

\begin{definition}[Generalized random dot product graph]
\label{def:grdpg}
Let $p \geq 1$ and $q \geq 0$ be integers. 
Define $I_{p,q}$ as the block diagonal matrix $I_{p,q} = \Bigl[\begin{smallmatrix} I_{p} & 0 \\ 0 & - I_{q} \end{smallmatrix} \Bigr]$ where $I_p$ and $I_q$ are the identity matrices of dimensions $p \times p$ and $q \times q$ respectively. 
Denote $d = p + q$ and let $\mathcal{X}$ be the subset of $\mathbb{R}^d$ such that, for any $x, y \in \mathcal{X}$, we have $x^\top I_{p,q} y \in [0, 1]$. 
Let $X = \big[ x_1 \mid \cdots \mid x_n \big]^\top$ be an $n \times d$ matrix with rows $x_i \in \mathcal{X}$. 
A graph $G$ with adjacency matrix $A$ is said to be a generalized random dot product graph with latent positions $X$, sparsity parameter $\rho_n$, and signature $(p, q)$ if $A \sim \BG(P)$ where the edge probability matrix $P$ is given by $P = \rho_n X I_{p,q} X^\top$, i.e., the entries of $P$ are of the form $P_{ij} = \rho_n x_i^\top I_{p,q} x_j$.

@@

\begin{example}
Suppose that latent vectors for an RDPG are drawn from the curve 
$x(t) = \big[ t^2, 2 t (t-1) \big]^\top$, $t \in [0, 1]$. 
Let $x_i = x(t_i)$. 
Then for each $i < j$, the edges are drawn as $A_{ij} \indep \Bernoulli(x_i^\top x_j)$. 
If after observing $A$, we wish to estimate the curve, the ASE provides an estimator in which the estimated latent vectors converge to the curve, up to an orthogonal transformation (figure \ref{fig:ase-example}).

@@

\begin{example}
In the $\ER$ model, the edge probability matrix has a fixed value $P_{ij} \equiv p \in [0, 1]$. 
This corresponds to a one-dimensional RDPG with a latent structure $X = \big[ \sqrt{p},  \cdots, \sqrt{p} \big]^\top$. 
In another view, the RDPG with latent structure $X = \big[ \sqrt{p}, \cdots, \sqrt{p} \big]^\top$ induces an $\ER$ model in which the edge probabilities have a fixed value.

Suppose that we have RDPG in which the latent space is in $\mathbb{R}^2$ and the latent vectors are drawn uniformly from the quarter unit circle defined by $g(t) = \big[ \cos \frac{\pi}{2} t, \sin \frac{\pi}{2} t \big]^\top$ for $t \in [0, 1]$. 
Then it can be shown that instead of a fixed $P_{ij} = p$, this RDPG induces a model in which the edge probabilities follow a distribution with density $f(p) = \frac{2}{\pi - 2} \Big( \frac{1}{1 - p^2} - 1 \Big)$. 
\end{example}

@@

\begin{enumerate}
\item Draw labels $z_1, ..., z_n \iid \Multinomial(\alpha_1, ..., \alpha_K)$.
\item Draw latent vectors by first taking each $t_i \indep F_{z_i}$ and then computing each $x_i = g_{z_i}(t_i)$. 
\item Compile the latent vectors into data matrix $X = \big[ x_1 \mid \cdots \mid x_n \big]^\top$ and define the adjacency matrix as $A \sim \GRDPG_{p,q}(X; \rho_n)$. 
\end{enumerate}
\end{definition}

@@

\begin{example}
\label{ex:separated-curves}
Define two one-dimensional manifolds in $\mathbb{R}^2$ by $f_1(t) = \big[ \cos(\frac{\pi}{3} t), \sin(\frac{\pi}{3} t) \big]^\top$ and $f_2(t) = \big[ 1 - \cos(\frac{\pi}{3} t), 1 - \sin(\frac{\pi}{3} t) \big]^\top$.
Draw $t_1, ..., t_n \iid \Uniform(0, 1)$ and $z_1, ..., z_n \iid \Multinomial(\frac{1}{2}, \frac{1}{2})$, and compute latent vectors $x_i = f_{z_i}(t_i)$, which are collected in data matrix $X = \big[ x_1 \mid \cdots \mid x_n \big]^\top$. 
Finally, let $A \sim \RDPG(X)$. Figure~\ref{fig:two-curves} shows the latent configuration drawn from this latent distribution, a random dot product graph drawn from the latent configuration, and the ASE of the graph. 
Although the community structure is not obvious from the graph, the embedding shows a clear separation between the two communities.

@@

\begin{theorem}[Community detection for the GRDPG for which the communities come from nonintersecting manifolds]
\label{theorem:nonintersect-grdpg}
Let $x_1, ..., x_n$ be points sampled on $K$ compact, connected manifolds $\mathcal{M}_1, ..., \mathcal{M}_K \subset \mathbb{R}^d$ each with probability measures $F_1, ..., F_K$, and the manifolds are separated by distance at least $\delta = \min\limits_{k \neq \ell} \min\limits_{x_i \in \mathcal{M}_k, x_j \in \mathcal{M}_\ell} \| x_i - x_j \| > 0$. 
Let $X = \big[ x_1 \mid \cdots \mid x_n \big]^\top$ and $A \sim \GRDPG_{p,q}(X; \rho_n)$ for some $p, q \in \mathbb{N}_0$ such that $p + q = d$ and sparsity parameter $\rho_n$ that satisfies $n \rho_n = \omega(\log^c n)$ for some $c > 1$. 
Define $A_n(\eta)$ as the event that an $\eta$-neighborhood graph constructed from the ASE of $A$ consists of exactly $K$ disconnected subgraphs that map exactly to each manifold. 
Then for some $C > 0$ and any $\eta \in (0, C \delta)$, $\lim\limits_{n \to \infty} P(A_n(\eta)) = 1$. 
\end{theorem}

@@

1. Draw $t_1, ..., t_n \iid F$ for probability distribution $F$ with support $[0, 1]$.
2. Draw $z_1, ..., z_n \iid \Multinomial(\alpha_1, ..., \alpha_K)$, the community labels.
3. Let each $x_i = g_{z_i}(t_i)$ be the latent vector for vertex $v_i$, and collect the latent vectors into matrix $X = \big[ x_1 \mid \cdots \mid x_n \big]^\top$.
4. Draw $A \sim \GRDPG_{p,q}(X; \rho_n)$.

\begin{example}
\label{ex:intersect-curves}
Suppose we have a two-dimensional RDPG in which latent vectors lie on one of two curves defined by the parameterizations $g_1(t) = \big[ t^2, 2 t (1-t) \big]^\top$ and $g_2(t) = \big[ 2 t (1-t), (1-t)^2 \big]^\top$. 
This can be described as a two-community MBM in which the manifolds are both one-dimensional. 
For simplicity, we restrict $F_1 = F_2 = \Uniform(0, 1)$. 
$g_1$ and $g_2$ intersect at two points, $t = 0$ and $t = 2/3$, which in the latent space correspond to $(0, 0)$ and $(4/9, 4/9)$. 
Clearly, single-linkage or $\eta$-neighborhood clustering does not work for this setting, both in the original latent space and in the ASE. 
The community structure is also not clear from the adjacency matrix. 
However, there still is a clear pattern in the ASE that reveals the community structure (figure \ref{fig:intersect-curves-example}).

@@

\begin{example}
\label{ex:intersect-curves-fit}
Figure \ref{fig:intersect-curves-2} shows one result of applying algorithm \ref{alg:kcurves} on example \ref{ex:intersect-curves}. 
In this implementation, $g_1$ and $g_2$ are assumed to be Bezier curves in which $p_1^{(0)} = p_2^{(0)} = \big[ 0, 0 \big]^\top$, i.e., the curves meet at the origin. 
Here, an accuracy of 87.5\% is achieved.

@@

We performed two simulation experiments for algorithm \ref{alg:kcurves}.
In the first simulation setup, we again turn to examples \ref{ex:intersect-curves} and \ref{ex:intersect-curves-fit}. 
The setup is as follows:

1. Draw $z_1, ..., z_n \iid \Multinomial(1/2, 1/2)$.
2. Draw $t_1, ..., t_n \iid \Uniform(0, 1)$. 
3. Let each $x_i = g_{z_i}(t_i)$ where  $g_1(t) = \big[ t^2, 2 t (1-t) \big]^\top$ and $g_2(t) = \big[ 2 t (1-t), (1-t)^2 \big]^\top$. 
Collect the latent vectors into matrix $X = \big[ x_1 \mid \cdots \mid x_n \big]^\top$. 
4. Draw $A \sim \RDPG(X; \rho_n)$ (for these simulations we fix $\rho_n \equiv 1$).

For this simulation study, we set $n = 128, 256, 512, 1024, 2048$, and we start with completely random initialization as well as initializations in which we know 4 or 8 labels from each community. 
For each setting, we simulate 50 replicates. 
Since the curves meet at the origin, we force the fitted Bezier curves to meet at the origin as well by fixing $p_1^{(0)} = p_2^{(0)} = (0, 0)$. 
Figure \ref{fig:sim-curves} shows the median and interquartile ranges for the error rates for each $n$ and *a priori* known labels per community.
The simulation results suggest that for sufficient $n$, we do not benefit from having more known labels, but having a few results in greatly improved performance.

```{r sim-curves, fig.height = 3, fig.width = 8, fig.cap = 'Median and IQR error rates over 50 replicates in the simulation setting for $K = 2$ curves.', cache = TRUE}
n.vec <- 2 ^ c(7, 8, 9, 10, 11)
nsamp.vec <- c(0, 4, 8)

clustering.df <- readr::read_csv('~/dev/manifold-block-models/simulations/balanced-2/balanced.csv')

@@

In the second simulation, we extended the first simulation setup based on examples \ref{ex:intersect-curves} and \ref{ex:intersect-curves-fit} into three dimensions and three curves. 
The setup is as follows:

1. Draw $z_1, ..., z_n \iid \Multinomial(1/3, 1/3, 1/3)$.
2. Draw $t_1, ..., t_n \iid \Uniform(0, 1)$. 
3. Let each $x_i = g_{z_i}(t_i)$ where
    i. $g_1(t) = \big[ 2 t (t-1), t^2, 0 \big]^\top$
    ii. $g_2(t) = \big[ 0, t^2, 2 t (t-1) \big]^\top$
    iii. $g_3(t) = \big[2 t (t-1), 0, t^2 \big]^\top$.
4. Collect the latent vectors into matrix $X = \big[ x_1 \mid \cdots \mid x_n \big]^\top$.
5. Draw $A \sim \RDPG(X; \rho_n)$ (for these simulations we fix $\rho_n \equiv 1$).

As in the previous setup, we set $n = 128, 256, 512, 1024, 2048$, and we start with completely random initializations as well as initializations in which we know $4$ or $8$ labels from each community. 
For each setting, we simulate 50 replicates. 
Since the curves meet at the origin, we force the fitted Bezier curves to meet at the origin as well by fixing $p_1^{(k)} = (0, 0, 0)$ for each $k$. 
Figure \ref{fig:sim-curves-3} shows the median and interquartile ranges for the error rates for each $n$ and *a priori* known labels per community. 
As in the previous $K = 2$ setup, the results suggest that for sufficient $n$, we do not benefit from having more known labels, but having a few results in greatly improved performance. Unlike in the previous setup, random initialization fails to improve with greater $n$.

```{r sim-curves-3, fig.height = 3, fig.width = 8, fig.cap = 'Median and IQR error rates over 50 replicates in the simulation setup for $K = 3$ curves.', cache = TRUE}
n.vec <- 2 ^ c(7, 8, 9, 10, 11)
nsamp.vec <- c(0, 4, 8)

clustering.df <- readr::read_csv('~/dev/manifold-block-models/simulations/balanced-3/balanced-3.csv')

@@

plot.estimated.curves(Xhat, mbconnectome.out) + 
  scale_colour_brewer(palette = 'Set1') + 
  coord_fixed(.4) +
  labs(x = expression(x[1]), y = expression(x[2]))
```

@@

The statement of the theorem assumes that for each community $k$, there at least $R + 1$ points with known label $k$. 
For simplicity, suppose we have exactly $R + 1$ (if there are more, then discard the rest). 
Then for each community $k$, it is possible to fit a polynomial of order $R$ that goes through each of the $R + 1$ known points exactly (see appendix C), yielding $g(t; \hat{p}_k)$, where $\hat{p}_k$ are the fitted coefficients of the Bezier polynomial. 
Use the fitted Bezier curves to define $\delta_k(t) = \|Q g(t; \hat{p}_k) - g(t; p_k)\|$, $\delta_k = \max_t \delta_k(t)$, and $\delta = \max_k \delta_k$. 
Since $P(\epsilon < M n^{-1/2} \log^c n) \to 1$ for constants $M, c > 0$, 
$P(\delta_k < M n^{-1/2} \log^c n) \to 1$ for each $k$. 
Since $\epsilon_i$ describes the error between each embedding point and its corresponding latent vertex, and $\delta_k(t_i)$ describes the error between the fitted curve and its corresponding true curve in the latent space, 
$$L < \frac{1}{n} \sum_i \epsilon_i + \delta_{z_i}(t_i) < \frac{1}{n} n (\epsilon + \delta) = \epsilon + \delta.$$ 
Then as $n \to \infty$, 
$$P(L < 2 M n^{-1/2} \log^c n) > P(\epsilon + \delta < 2 M n^{-1/2} \log^c n) \to 1.$$
\end{proof}

@@

The curve fitting step in algorithm \ref{alg:kcurves} requires estimation of some $g: [0, 1] \to \mathbb{R}^d$ for each subset of points corresponding to each label. 
One choice of curve is the Bezier polynomial. 
The Bezier polynomial of order $R$ is defined by function $g$ with parameters $p$ such that
$$g(t; p) = \sum_{r=0}^R p^{(r)} \binom{R}{r} t^r (1-t)^{R-r}$$
where the Bezier coefficients $p^{(r)}$ are vectors in $\mathbb{R}^d$, and $p = \big[ p^{(0)} \mid \cdots \mid p^{(R)} \big]$.

The general idea behind Bezier curve fitting to noisy data is as follows: 
We wish to minimize the squared loss function 
\begin{equation}
\label{eq:bezier-ls}
L(p, t; X) = \sum_i \|x_i - g(t_i; p)\|^2.
\end{equation}
Suppose that we know the timepoints $t_1, ..., t_n$ which we wish to use to fit Bezier coefficients $p^{(0)}, ..., p^{(R)}$ to vectors $x_1, ..., x_n \in \mathbb{R}^d$ lying on or near a Bezier curve of order $R$. 
The squared loss then can be written as 
$$L(p; t, X) = \|X - T p\|^2,$$
where $p$ is defined as before, $T$ is an $n \times (R+1)$ matrix such that $T_{ir} = \binom{R}{r} t_i^r (1-t_i)^{R-r}$, and $X = \big[ x_1 \mid \cdots \mid x_n \big]$. 
Then the least squares estimates for the Bezier coefficients $p$ is simply the linear least squares estimate 
\begin{equation}
\label{eq:bezier-p-fit}
\hat{p} = (T^\top T)^{-1} T^\top X.
\end{equation}
